---
title: "Crayfish Exemple Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Crayfish Exemple Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
required <- c("edgeR")
evalChuncks <- all(unlist(lapply(required, function(pkg) requireNamespace(pkg, quietly = TRUE))))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = evalChuncks
)
```

```{r setup}
library(limma)
library(phyloDE)
```

## Data

In this tutorial, we re-analyse the Crayfish dataset from
Stern & Crandal (2018)[^1], re-analysed in Bastide et al. (2023)[^2],
and available in the `crayfish` dataset.

```{r}
data(crayfish)
```

The dataset contains a count RNA-Seq matrix,
associated gene lengths,
a phylogenetic tree of the crayfish species included in the study,
and a data-frame specifying whether a species is blind (0) or sighted (1).

```{r, fig.show='hold', fig.height=4, fig.width=5}
library(ape)
cond_species <- crayfish$sights$sights[match(crayfish$tree$tip.label, crayfish$sights$species)]
plot(crayfish$tree, tip.color = hcl.colors(3)[as.numeric(cond_species)])
```


## Normalization of the data

In order to apply `phyloDE`, we need to normalize the data.
Here, we use TPM, with normalizing factors computed using the TMM method.

```{r}
nf <- edgeR::calcNormFactors(crayfish$counts / crayfish$lengths, method = "TMM")
norm_data <- lengthNormalizeRNASeq(crayfish$counts, 
                                   crayfish$lengths, 
                                   nf, 
                                   lengthNormalization = "TPM", 
                                   dataTransformation = "log2")
```


## Design matrix

We then construct the design matrix, to analyse the effect of sight
on gene expression.

```{r}
design <- model.matrix(~ sights, model.frame(crayfish$sights))
```

## Heatmap matrix

We can plot the data with a heatmap, using the phylogenetic
structure on the columns, and a standard clustering on the rows.
```{r, fig.show='hold', fig.height=7, fig.width=7}
phyHeatmap(norm_data, design, coef = 2, crayfish$tree, margins = c(7, 3))
```

## Fit with `phyloDE`

Finally, we apply `phylolmFit`, that has a syntax similar to 
`limma` function `lmFit`.
As an additional argument, it takes the phylogenetic tree.

To keep compilation time low, we only analyse the first 500 genes here,
but in an analysis users should use the full dataset, and
parallelize the computations using the `ncores` argument.



```{r}
pfit <- phylolmFit(norm_data[1:500, ], design,
                   phy = crayfish$tree,
                   ncores = 1)
```


## Empirical Bayes analysis

Just as for a `limma` object, we can apply the `eBayes` procedure
to the results.

Keep in mind that because we only analysed the first 500 genes, the results
cannot to be interpreted directly, and some plots might look different when
analyzing the full dataset.

We can check whether there is a trend in the data by plotting the mean-variance trend.
```{r, fig.show='hold', fig.height=4, fig.width=5}
mean_genes <- pfit$Amean
sd_genes <- sqrt(pfit$sigma)
lfit <- lowess(mean_genes, sd_genes, f = 0.5)
plot(mean_genes, sd_genes,
     xlab = "log2(TPM)", ylab = "Sqrt( standard deviation )",
     pch = 16, cex = 0.25)
title("Mean-variance trend")
lines(lfit, col = "red")
```

We then apply `eBayes`, here with a trend.

```{r}
pfit <- eBayes(pfit, trend = TRUE)
```

And then extract the most significant genes with `topTable`.
```{r}
topTable(pfit, coef = 2)
```

## Standard Diagnostics

As for a standard `limma` analysis, we can perform some diagnostic plots.

* Plot of p-values: they must be uniformly distributed under the null hypothesis.
```{r, fig.show='hold', fig.height=4, fig.width=5}
hist(pfit$p.value)
```

* MA Plot: points must be centered in zero, showing no biais.
```{r, fig.show='hold', fig.height=4, fig.width=5}
plotMD(pfit)
abline(h = 0, col = "red", lwd = 3)
```

* Volcano plot:
```{r, fig.show='hold', fig.height=4, fig.width=5}
volcanoplot(pfit, coef = 2, highlight = 10)
```

* Heatmap restricted to top $10$ genes:
```{r, fig.show='hold', fig.height=7, fig.width=7}
topGenes <- rownames(topTable(pfit, coef = 2))
phyHeatmap(norm_data[match(topGenes, rownames(norm_data)), ], design, coef = 2, crayfish$tree, margins = c(7, 6))
```

## Parameters of the OU

Some diagnostic plots are specific to `phyloDE`,
that fits a OU process per gene, and then regularize the parameters
(by taking the trimmed means of the tanh values).

Function `getParameters` extract the regularized parameters.
```{r}
getParameters(pfit)
```
The parameters are the normalized `lambda` and `rho` parameters:

* `lambda` is the ratio of noise attributed to the phylogeny versus the total noise.
    - Values close to 0 mean that the tree has little influence (independent errors only).
    - Values close to 1 mean that there the tree explains all the variance (no independent errors).

* `rho` is the percent decrease in
trait variance caused by the OU as compared to the variance expected under under BM,
see Cornuault (2022)[^3].
    - Values close to 0 mean that process looks like a BM (selection strength alpha is close to zero).
    - Values close to 1 mean that the tree is close to a star tree (selection strength alpha is large).

The gene-specific and regularized parameters can be plotted with
function `plotParameters`.

```{r, fig.show='hold', fig.height=4, fig.width=7}
plotParameters(pfit)
```
Histograms are histograms over all the fits on all genes,
while dashed lines represent regularized parameters.
Modes close to the 0 and 1 bounds can be expected,
especially for small parameters.



[^1]: Stern, D. B. and Crandall, K. A. (2018), 
'The Evolution of Gene Expression Underlying Vision Loss in Cave Animals', 
Molecular Biology and Evolution 35(8), 2005--2014.

[^2]: Bastide, P., Soneson, C., Stern, D. B., Lespinet, O. and Gallopin, M. (2023), 
'A Phylogenetic Framework to Simulate Synthetic Interspecies RNA-Seq Data', 
Molecular Biology and Evolution 40(1), msac269.

[^3]: Cornuault, J. (2022), 
'Bayesian Analyses of Comparative Data with the Ornstein–Uhlenbeck Model: Potential Pitfalls', 
Systematic Biology 71(6), 1524--1540.
